---
title: "Data Analyst ND P4"
author: "Wilfried Hoge"
date: "17. Oktober 2015"
output: html_document
---

# Taxi Exploration by Wilfried Hoge

The Taxi data used in this exploration was downloaded from 
http://www.andresmh.com/nyctaxitrips/ and joined with weather data from 
https://weatherspark.com/. This data set is very large and only a small sample 
is used here. This sample contains 1% of the data for 4 month 
(January to April 2013).

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}

library(ggplot2)
library(dplyr)
Sys.setlocale("LC_TIME", "C")

```


## General Statistics

```{r echo=FALSE, error=FALSE, warning=FALSE}

# my working directory
setwd("~/Dropbox/udacity")

# read taxi data from csv
taxi = read.csv("taxi.csv", 
                colClasses=(c("factor","factor","character","character",
                              "factor","integer","numeric","numeric","numeric",
                              "numeric","numeric","factor","numeric","numeric",
                              "numeric","numeric","numeric","numeric","factor",
                              "factor","factor","factor","numeric","numeric")))

# build some new variables useful for analysis
taxi$pickup.date = as.Date(taxi$pickup.datetime)
taxi$weekday = factor(weekdays(taxi$pickup.date),
                      level=c("Monday", "Tuesday", "Wednesday", "Thursday", 
                              "Friday", "Saturday", "Sunday"))
taxi$weekend = factor(ifelse(taxi$weekday %in% c("Saturday","Sunday"), 
                             "yes", "no"))
taxi$rain = factor(ifelse(taxi$Precipitation>0, "yes", "no"))
taxi$Hour = factor(taxi$Hour, levels=c(0:23))

```

```{r}
# show the structure of the taxi data
str(taxi)

# show the factors
levels(taxi$Vendor)
levels(taxi$passenger.count)
levels(taxi$Hour)
levels(taxi$weekday)
levels(taxi$weekend)
levels(taxi$rain)

# remove NAs from the data set
taxi = na.omit(taxi)

# a summary of the data
summary(taxi)

```

Most taxi trips are just short. The mean trip distance is below 3 miles. 
The mean passenger count is 1.7. The median tip is just $1 but the maximum 
is $110. The mean trip fare is $14.33 and the maximum is $356.

## Simple plots

Looking at the taxi rides per day shows a distribution between 400 and 600. 
The monday has the lowest number of rides with a median of 450. The highest 
median of taxi rides is on Friday and Saturday with a higher variance on 
Saturday.
There are not many taxi rides when it is raining.

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=taxi, aes(x=pickup.date)) + geom_histogram(binwidth=1) +
  ggtitle("Number of taxi rides per day") + xlab("Date")

taxi.per.day = taxi %>% 
  group_by(pickup.date) %>% 
  summarise(count=n(), fare.mean=mean(fare.amount),
            fare.median=median(fare.amount),
            weekday=first(weekday))

ggplot(data=taxi.per.day, aes(x=weekday,y=count)) + geom_boxplot() + 
  ggtitle("Number of taxi rides per weekday")

ggplot(data=taxi, aes(x=rain)) +  geom_histogram() +
  ggtitle("Number of taxi rides by rain") + xlab("Is it raining?")

```

## Looking at fares

The distribution of fares over days shows that most trips are below $25 
(the mean is $14.33). But a line above $50 stands out that should be 
investiged in more detail.

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=taxi, aes(x=pickup.date,y=fare.amount)) + 
  geom_point(alpha=0.1, color="orange", position="jitter") + ylim(0,100) + 
  geom_line(data=taxi.per.day, (aes(x=pickup.date, 
                                    y=fare.mean, color="mean"))) +
  geom_line(data=taxi.per.day, (aes(x=pickup.date, 
                                    y=fare.median, color="median"))) +
  ggtitle("Trip fare per day") + xlab("Day") + ylab("Fare") + 
  scale_colour_manual("Fare", values=c("mean"="blue",
                                       "median"="red"))

```

Looking at trip fares above $50 we can see that the rate of $52 is very 
frequent.

```{r}

# subset of taxi trips with fare > 50$
taxi50 = taxi[taxi$fare.amount > 50,]
table(taxi50$fare.amount)

```

This high frequency of this special rate looks like a fixed price offering 
from/to the airport. To verify this, the geo coordinates are checked to look 
at the start and end point of the trips.

```{r echo=FALSE, error=FALSE, warning=FALSE}

library(ggmap)
nyc.map = get_map(location = c(lon = -73.9, lat = 40.7), zoom = 11, 
                  maptype = 'roadmap')
nyc.MAP = ggmap(nyc.map)

# subset of taxi trips that start or end beyond the -73.9 lon line 
# (this is not exactly selecting the airport but JFK airport is on on side and 
#  Manhattan is on the other side of the line)
taxi50.from.ap = taxi50[taxi50$pickup.longitude > -73.9 & 
                          taxi50$trip.distance > 15,]
taxi50.to.ap   = taxi50[taxi50$dropoff.longitude > -73.9 & 
                          taxi50$trip.distance > 15,]

nyc.MAP + 
  geom_point(data=taxi50.from.ap, aes(x=pickup.longitude,y=pickup.latitude), 
             alpha=.1, color="red") + 
  geom_point(data=taxi50.from.ap, aes(x=dropoff.longitude,y=dropoff.latitude), 
             alpha=.1, color="blue") +
  ggtitle("Taxi rides from Airport(red) to NCY(blue)")

nyc.MAP + 
  geom_point(data=taxi50.to.ap, aes(x=pickup.longitude,y=pickup.latitude), 
             alpha=.1, color="red") + 
  geom_point(data=taxi50.to.ap, aes(x=dropoff.longitude,y=dropoff.latitude), 
             alpha=.1, color="blue") +
  ggtitle("Taxi rides from NYC(red) to Airport(blue)")

```

The maps show that the special price trips start/end at the airport and 
end/start in Manhattan.

## Look at trips to Airport in more detail

Arriving the airport in time is important. Therefore we look at the time it 
takes to ride from the city of Manhattan to the airport.

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=taxi50.to.ap, aes(x=Hour, y=trip.time)) + 
  geom_jitter(alpha=1, aes(color=weekend)) + 
  geom_smooth(method="loess", aes(group=1)) +
  ggtitle("Trip time to Aiport over hour of day") +
  ylab("Trip time")

```

The trip time to the airport is dependent on the hour of the day. Around 4pm 
in the afternoon it takes much longer than around 5am. The trip times on the 
weekend are mostly below the fitting line, showing that it is easier to drive 
to the airport on weekends.

## Trying to predict the trip time to the airport

As it is important to know how long it will take to drive to the airport a 
prediction model is created. The model is not giving a good prediction 
(R squared value below 0.35). The reason is the small dataset and the large 
variance of the trip times on some hours of the day. The graph below shows 
the distribution of trip times to the airport, the predicted trip times 
(in blue) and the upper limit of the 99% confidence interval (in red).

```{r}

lm1 = lm(trip.time ~ Hour, data=taxi50.to.ap)
summary(lm1)

# build a test data set to predict (5am to 8pm on workdays)
test=data.frame(Hour=factor(c(5:20),levels=c(5:20)), weekend="no")

# do prediction on test data
pred = predict(lm1, test, interval = c("confidence"), level = 0.99)
pred = data.frame(pred)
test$fit = pred$fit
test$upr = pred$upr

# show predicted data and upper limit for CI
test

```

```{r echo=FALSE, error=FALSE, warning=FALSE}

# Plot data with predicted values and CI
ggplot(data=subset(taxi50.to.ap, trip.distance > 15), 
       aes(x=Hour, y=trip.time)) + 
  geom_jitter(alpha=.8, color="orange") + 
  ggtitle("Trip time to Aiport over hour of day") +
  ylab("Trip time") + ylim(1000,4000) +
  geom_line(data=test, aes(x=Hour, y=fit, group=1, color="Predicted")) +
  geom_line(data=test, aes(x=Hour, y=upr, group=1, color="99% Confidence")) +
  scale_colour_manual("Prediction", values=c("Predicted"="blue",
                                             "99% Confidence"="red"))

```

## Looking at fare per mile and speed

The taxi fare per mile and speed are distributed as expected. The mean fare 
per mile is 5.87$ and the mean speed is 13.3 miles per hour. The histogram for 
the fair per mile has a long tail. Cutting of the upper 0.01% of the data 
gives a better overview of the distribution.

```{r}

# Subselecting taxi trips with distance and trip time > 0
taxi0 = taxi[taxi$trip.distance>0 & taxi$trip.time>0,]

# calculate new varaibles for fare.per.mile and speed 
taxi0$fare.per.mile = taxi0$fare.amount/taxi0$trip.distance
taxi0$speed = taxi0$trip.distance/taxi0$trip.time*3600

# omit trips with speed too high
taxi0 = taxi0[taxi0$speed < 100,]

# distribution of new variables
summary(taxi0$fare.per.mile)
summary(taxi0$speed)

```

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=taxi0, aes(x=speed)) + geom_histogram(binwidth=1) +
  ggtitle("Distribution of speed") + 
  scale_x_continuous(breaks=c(0,5,10,15,20,30,40,60), limit=c(0,60))

ggplot(data=taxi0, aes(x=fare.per.mile)) + geom_histogram(binwidth=1) +
  xlab("Fare per mile") + ggtitle("Distribution of fare per mile")

ggplot(data=taxi0, aes(x=fare.per.mile)) + geom_histogram(binwidth=.5) +
  coord_cartesian(xlim=c(0,quantile(taxi0$fare.per.mile,.99))) +
  xlab("Fare per mile") + ggtitle("Distribution of fare per mile (lower 99%)")
  
```

Comparing the fair per mile for workdays and weekends shows that the fare is 
slightly higher for workdays. This might be due to the higher traffic compared 
to the weekend.

Looking at the speed per weekday you can see that the median speed for Sunday 
is the highest and Friday has the lowest.

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=taxi0, aes(x=weekend, y=fare.per.mile)) + geom_boxplot() +
  coord_cartesian(ylim=c(0,quantile(taxi0$fare.per.mile,.95))) +
  ggtitle("Comparison of taxi fare per mile for workdays vs. weekend") +
  xlab("Is weekend?") + ylab("Fare per mile")

ggplot(data=taxi0, aes(x=weekday, y=speed)) + geom_boxplot() +
  coord_cartesian(ylim=c(0,quantile(taxi0$speed,.95))) +
  ggtitle("Comparison of Speed per workday")

```

Comparing the speed over the hour of day shows smaller number of taxi rides in 
the early morning hours and the slowest speeds in the early afternoon. The 
smoothed line for the mean of the speed visualizes the dependency of speed 
from hour of day.

Seperating the speed by weekend or workdays in a boxplot shows that nightly 
trips on weekends (Saturday or Sunday morning) are slower than on worksdays.
During daytime the speeds are faster on weekends than on workdays.

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=taxi0, aes(x=Hour, y=speed)) + geom_jitter(alpha=0.1, 
                                                       color="orange") +
  coord_cartesian(ylim=c(0,quantile(taxi0$speed,.95))) +
  ggtitle("Comparison of Speed per hour of day")

ggplot(data=taxi0, aes(x=Hour, y=speed)) + geom_jitter(alpha=0.1, 
                                                       color="orange") +
  coord_cartesian(ylim=c(0,quantile(taxi0$speed,.95))) +
  geom_smooth(aes(group=1)) +
  ggtitle("Comparison of Speed per hour of day")

ggplot(data=taxi0, aes(x=Hour, y=speed)) + 
  geom_boxplot(aes(color=weekend),outlier.colour = "grey") +
  coord_cartesian(ylim=c(0,quantile(taxi0$speed,.95))) +
  ggtitle("Comparison of Speed per hour of day")

```

## Looking at the influence of rain

Before comparing taxi rides on weather data three plots visualize the weather 
data. The temperature is shown as points and tiles whereas the rain is shown 
as tiles only. There are some "holes", where no weather data is available, 
because no taxi trips are in the sample at that point in time.

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=taxi, aes(x=pickup.date, y=Temperature)) + 
  geom_jitter(alpha=.1, color="orange") + geom_smooth() +
  ggtitle("Temperature") + xlab("Date")

ggplot(data=taxi, aes(x=pickup.date, y=Hour, fill=Precipitation)) + 
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("white", "red"))(5)) + 
  ggtitle("Rain per day and hour") + xlab("Date")

ggplot(data=taxi, aes(x=pickup.date, y=Hour, fill=Temperature)) + geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(5)) + 
  ggtitle("Temperature per day and hour") + xlab("Date")

```

Comparing the speed of the taxis by rain visually gives no information. 
Looking at the mean and median values, a small difference visible. Taxis are 
slower and the fares are higher, when it is raining. Also, the trip distance 
is shorter when it is raining, pointing at the fact that a taxi might be used 
even for smaller distances.

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=taxi0, aes(x=Hour, y=speed)) + geom_jitter(alpha=0.5, 
                                                       aes(color=rain)) +
  coord_cartesian(ylim=c(0,quantile(taxi0$speed,.95))) +
  ggtitle("Comparison of Speed per hour of day")

```

```{r}
by(taxi0[,c("rain","speed","trip.distance","tip.amount")], taxi0$rain, summary)
```

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=taxi0, aes(x=rain, y=speed)) + geom_boxplot() +
  coord_cartesian(ylim=c(0,quantile(taxi0$speed,.95))) +
  ggtitle("Comparison of Speed by Rain") + xlab("Is it raining?")

ggplot(data=taxi0, aes(x=rain, y=fare.per.mile)) + geom_boxplot() +
  coord_cartesian(ylim=c(0,quantile(taxi0$fare.per.mile,.95))) +
  ggtitle("Comparison of Fare per mile by Rain") + 
  xlab("Is it raining?") + ylab("Fare per mile")

ggplot(data=taxi0, aes(x=rain, y=trip.distance)) + geom_boxplot() +
  coord_cartesian(ylim=c(0,quantile(taxi0$trip.distance,.90))) +
  ggtitle("Comparison of Trip distance by Rain") + 
  xlab("Is it raining?") + ylab("Trip distance")


```

## Looking at tips

The distribution of tips over trip distance shows that many people don't give 
tips at all. For longer distances, the tips rises a little bit but on average 
less than $1 a mile.

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=subset(taxi0,tip.amount<10), 
       aes(y=trip.distance, x=factor(tip.amount))) + 
  geom_jitter(alpha=.1, color="orange") +
  scale_y_continuous(breaks=(c(0:10,15,20)), limit=c(0,20)) +
  geom_boxplot(aes(y=trip.distance,x=factor(tip.amount)),
               color=I("#56B4E9"), 
               outlier.colour="grey", outlier.size=0,
               alpha=0) + 
  coord_flip() +
  ggtitle("Distribution of tips over trip distance") + 
  xlab("Tip") + ylab("Trip distance")
```

Even more interesting is the fact that the mean tip is less for 3 or 4 
passengers in a taxi than it is for 1, 2 or 5, and 6 passengers. I have no 
idea what might be the cause of this fact.

```{r echo=FALSE, error=FALSE, warning=FALSE}

by(taxi0[,c("tip.amount","passenger.count")], taxi0$passenger.count, summary)

```

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=taxi0, aes(x=passenger.count, y=tip.amount)) + 
  geom_jitter(alpha=.2, color="orange") +
  geom_line(stat="summary", fun.y=mean, aes(group=1, color="mean")) + 
  coord_cartesian(ylim=c(0,5)) +
  ggtitle("Comparison of Tip and Passenger count") + 
  xlab("Passenger count") + ylab("Tip [$]") +
  scale_colour_manual("", values=c("mean"="blue"))

```

# Final Plots and Summary

## Plot 1

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=taxi, aes(x=pickup.date,y=fare.amount)) + 
  geom_point(alpha=0.1, color="orange", position="jitter") + ylim(0,100) + 
  geom_line(data=taxi.per.day, (aes(x=pickup.date, 
                                    y=fare.mean, color="mean"))) +
  geom_line(data=taxi.per.day, (aes(x=pickup.date, 
                                    y=fare.median, color="median"))) +
  ggtitle("Trip fare per day") + xlab("Day") + ylab("Fare") + 
  scale_colour_manual("Fare in $", values=c("mean"="blue",
                                            "median"="red"))

```

## Description 1

The distribution of fares over days shows that most trips are below $25 
(the mean is $14.33). But a line above $50 stands out. A special fix price 
offering to get from the city to the airport and backward might be the reason.

## Plot 2

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=taxi50.to.ap, aes(x=Hour, y=trip.time)) + 
  geom_jitter(alpha=1, aes(color=weekend)) + 
  geom_smooth(method="loess", aes(group=1)) +
  ggtitle("Trip time to Aiport over hour of day") +
  ylab("Trip Time [seconds]")

```

## Description 2

The trip time to the airport is dependent on the hour of the day. Around 4pm 
in the afternoon it takes much longer than around 5am. The trip times on the 
weekend are mostly below the fitting line, showing that it is easier to drive 
to the airport on weekends.

## Plot 3

```{r echo=FALSE, error=FALSE, warning=FALSE}

ggplot(data=subset(taxi0,tip.amount<10), 
       aes(y=trip.distance, x=factor(tip.amount))) + 
  geom_jitter(alpha=.1, color="orange") +
  scale_y_continuous(breaks=(c(0:10,15,20)), limit=c(0,20)) +
  geom_boxplot(aes(y=trip.distance,x=factor(tip.amount)),
               color=I("#56B4E9"), 
               outlier.colour="grey", outlier.size=0,
               alpha=0) + 
  coord_flip() +
  ggtitle("Distribution of tips over trip distance") + 
  xlab("Tip [$]") + ylab("Trip Distance [miles]")

```

## Description 3

The distribution of tips over trip distance shows that many people don't give 
tips at all. For longer distances, the tips rises a little bit but on average 
less than $1 a mile.

# Reflection

The NYC Taxi data set contains a large amount of data. It has more than 10 
million taxi trips per month. In this investigation just a small subset of the 
trip data is analyzed (60.000 trips over 4 month). 

I started by looking at the distribution of taxi trips over date and weekdays 
but this gives not much information, as the distribution is almost even. 
Looking at the taxi fares showed a special price and it turned out that going 
from Manhattan to JFK airport or airport to Manhattan is the dominant taxi 
trip behind this special rate. Looking into trip times from Manhattan to JFS 
airport shows a strong dependency on hour of day. In the afternoon it takes 2x 
more time to arrive at the airport. I created a prediction model for this but 
the quality of this model was not very good based on the sample. It would be 
interesting to create and test the model with the full data set. Another 
interesting insight was that the mean tip amount is just around $1. As a 
German I have expected much higher tips. It is also interesting that with 
4 passengers in a taxi the mean of tips is the lowest.

There are numerous relationships that are not investigated here. E.g. the 
trip distance over time or the number of passengers compared by night or day 
could be also interesting. Looking at the full data set would give additional 
insights, that are not possible from the sample. E.g. it could be analyzed 
how many taxis are on the street in a given time frame or how many passengers 
each taxi has or how long taxis have to wait between customers.